The Community Land Model, version 5.1 \\ One-at-a-time Parameter Perturbation Ensemble

### 2023-05-16 ms/ms.tex
overleaf

Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within land components of Earth System Models.
Understanding how these variables respond directly and indirectly to CO$_2$ concentration, and how they themselves impact CO$_2$ concentrations, is at the center of climate change research.  
Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{esd-13-885-2022}.
Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, combined with increasingly comprehensive land surface modelling platforms, we might expect predictions to become more skillful across subsequent model generations.

While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations.
Within the day-to-day model development process, there is a greater emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration appears intractable \cite{dagon2020}.
Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g. \citeA{collier2018}.

One such fundamental model development step is parameter sensitivity testing and/or parameter optimization.
Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017}.
When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES).
They also form a basis for automated model calibration \cite{williamson2013history}.
Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes. 
In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, how it was produced, and its potential applications.

@@

\section{Experiment Description}
\label{methods}
\subsection{Model description}
\label{sect:md}
The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR). This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1). The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.

@@

To manage computational cost we leveraged the Matrix-CN spin-up mode recently implemented within CLM \cite{lu2020}. This new module utilizes a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time. Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset. This was designed to achieve sufficiently equilibrated model states, while minimizing computational time. This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter). Certain inferences about deep soil carbon are therefore subject to uncertainty due to spin-up concerns.

@@

\begin{table}[h]
\caption{Clustering inputs}
\centering
\begin{tabular}{l c c c c}
 \hline
 Climate forcing variables & Ecosystem state variables &Ecosystem flux variables \\
 \hline
 2m air temperature (TSA) & Leaf area index (TLAI) & Gross primary production (GPP) \\
Atmospheric rain (RAIN) & Ecosystem carbon (TOTECOSYSC) &Heterotrophic respiration (HR) \\
Atmospheric snow (SNOW) &  Ecosystem nitrogen (TOTECOSYSN) &Autotrophic respiration (AR) \\
2m specific humidity (Q2M) & Soil ice (TOTSOILICE) &Net biome production (NBP) \\
Solar radiation (FSDS) & Soil liquid water (TOTSOILLIQ) & Total liquid runoff (QRUNOFF) \\
& Snow cover fraction (FSNO) & Sensible heat  (FSH) \\
&&Latent heat (EFLX\_LH\_TOT ) \\
 \hline
 \end{tabular}
 \label{tab:sg}
 \end{table}

@@

\subsection{Parameters} 
\begin{landscape}
 \begin{table}[h]
 \caption{Some key parameters}
 \centering
 \begin{tabular}{l l c}
 \hline
  Parameter  & Description & Model Domain \\
 \hline
d\_max & Dry surface layer parameter & Sensible, latent heat and momentum fluxes \\
frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
fff & Decay factor for fractional saturated area & Hydrology \\
liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity (LUNA) \\
jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity (LUNA) \\
slatop & Specific leaf area at top of canopy & Photosynthetic capacity (LUNA) \\
wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity (LUNA) \\
kmax & Plant segment max conductance & Plant hydraulics \\
krmax & Root segment max conductance & Plant hydraulics \\
psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
nstem & Stem number & Biomass heat storage \\
lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
leafcn & Leaf C:N & Carbon and nitrogen allocation \\
leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
cpha & Activation energy for cp & Acclimation parameters \\
jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
kcha & Activation energy for kc & Acclimation parameters \\
lmrha & Activation energy for lmr & Acclimation parameters \\
lmrhd & Deactivation energy for lmr & Acclimation parameters \\
tpuha & Activation energy for tpu & Acclimation parameters \\
tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
vcmaxha & Activation energy for vcmax & Acclimation parameters \\
vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
 \hline
 \end{tabular}
 \end{table}
\end{landscape}

@@

A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.

@@

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
better the approximation oif output from the full grid can be acheived with the sparsegrid. Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sg}). We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
\begin{figure}[h]
\centering
\includegraphics[width=35pc]{../figs/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

@@

The PPE is likewise useful for determining which parameters influence which model processes and where. For example the parameters controlling leaf area index varies significantly by biome (Figure \ref{fig:lai}). Plant hydraulics were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome. The most influential parameters globally includes parameters from each of these three biomes. We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT})

@@

In this project, we have identified and perturbed 188 CLM parameters to create a large one-at-a-time parameter perturbation ensemble. This ensemble is useful for understanding parametric controls on CLM dynamics and has already been leveraged as a starting point for parameter optimization efforts. We made significant investments in software infrastructure to enable this experiment and reduce the burden of generating future PPEs.

@@

We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:violins}). That being said, the majority of parameter perturbations had very small effect, leading to output distributions strongly concentrated around the default simulation with long tails representing a small subset of parameters. In the case of gross primary production, only xx parameters had a parameter effect of more than 2\% under the control forcing scenario (Supp Figure \ref{supp:gpp}). As such, we would expect the the effective parameter space dimensionality for any given process would be significantly smaller than implied by the 188 parameters.

@@

A one-at-a-time PPE cannot capture parameter interactions, and our min/max sampling protocol precludes diagnosing non-linearities. As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty. The advantage of our dataset is that we can diagnose parameter effects without the uncertainty of an intermediating emulator or regression model. As such, it is easy to diagnose which parameters are most influential on a given process, which has been leveraged for parameter filtering. We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation. Now in addition to seeing the definition of a given parameter, and the relevant equations, you can see the magnitude and spatial patterns of its effects (e.g. Figure \ref{fig:panel}).

We have already begun a follow-on experiment that perturbs a subset of parameters with the latin hypercube sampling strategy to work towards calibration of leaf area index. By investing in the infrastructure underlying this first experiment, all of our subsequent experiments have required much less time and effort. We expect to continue to extend our efforts in this domain, as well as repeat these types of experiments with subsequent model releases.

>>>>

Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within land components of Earth System Models.
Understanding how these processes respond directly and indirectly to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is at the center of climate change research.  
Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{esd-13-885-2022}.
Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, combined with increasingly comprehensive land surface modelling platforms, we might expect predictions to become more skillful across subsequent model generations.

While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations (McNeall et al. 2016 ESD).
Within the model development process, there is a significant emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration appears intractable \cite{dagon2020}.
Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g. \citeA{collier2018}.

One such fundamental model development step is parameter sensitivity testing and/or parameter optimization (Qian et al. 2018 JGR Atmos).
Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017} (Balaji et al. 2022 PNAS).
When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES), (Peatier et al. 2022 GRL, Sanderson et al. 2008 JClim, Tett et al. 2022 JClim, Yamazaki et al. 2021 Clim Dyn).
PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013history} or Bayesian uncertainty quantification (Cleary et al. 2021 JCP).
Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes. 
In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, describe how it was produced, and survey some potential applications.

@@

\section{Experiment Description}
\label{methods}
\subsection{Model description}
\label{sect:md}
The Community Terrestrial Systems Model (CTSM) is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR). This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1). The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.

@@

To manage computational cost we leveraged the Matrix-CN spin-up mode recently implemented within CLM \cite{lu2020}. This new module utilizes a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time. Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset (described below). This protocol was designed to achieve sufficiently equilibrated model states, while minimizing computational time. This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth). Certain inferences about deep soil carbon are therefore subject to uncertainty due to spin-up concerns.

@@

\begin{table}[h]
\caption{Clustering inputs categorized into three groups, with the CLM variable name in parentheses}
\centering
\begin{tabular}{l c c c c}
 \hline
 Climate forcing variables & Ecosystem state variables &Ecosystem flux variables \\
 \hline
 2m air temperature (TSA) & Leaf area index (TLAI) & Gross primary production (GPP) \\
Atmospheric rain (RAIN) & Ecosystem carbon (TOTECOSYSC) &Heterotrophic respiration (HR) \\
Atmospheric snow (SNOW) &  Ecosystem nitrogen (TOTECOSYSN) &Autotrophic respiration (AR) \\
2m specific humidity (Q2M) & Soil ice (TOTSOILICE) &Net biome production (NBP) \\
Solar radiation (FSDS) & Soil liquid water (TOTSOILLIQ) & Total liquid runoff (QRUNOFF) \\
& Snow cover fraction (FSNO) & Sensible heat  (FSH) \\
&&Latent heat (EFLX\_LH\_TOT ) \\
 \hline
 \end{tabular}
 \label{tab:sg}
 \end{table}

@@

\subsection{Parameters} 
\begin{landscape}
 \begin{table}[h]
 \caption{Some key parameters}
 \centering
 \begin{tabular}{l l c}
 \hline
  Parameter  & Description & Model Domain \\
 \hline
d\_max & Dry surface layer (DSL) parameter & Sensible, latent heat and momentum fluxes \\
frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
fff & Decay factor for fractional saturated area & Hydrology \\
liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity (LUNA) \\
jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity (LUNA) \\
slatop & Specific leaf area at top of canopy & Photosynthetic capacity (LUNA) \\
wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity (LUNA) \\
kmax & Plant segment max conductance & Plant hydraulics \\
krmax & Root segment max conductance & Plant hydraulics \\
psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
nstem & Stem number & Biomass heat storage \\
lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
leafcn & Leaf C:N & Carbon and nitrogen allocation \\
leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
cpha & Activation energy for cp & Acclimation parameters \\
jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
kcha & Activation energy for kc & Acclimation parameters \\
lmrha & Activation energy for lmr & Acclimation parameters \\
lmrhd & Deactivation energy for lmr & Acclimation parameters \\
tpuha & Activation energy for tpu & Acclimation parameters \\
tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
vcmaxha & Activation energy for vcmax & Acclimation parameters \\
vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
 \hline
 \end{tabular}
 \end{table}
\end{landscape}

@@

A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.

@@

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
better the approximation oif output from the full grid can be acheived with the sparsegrid. Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sg}). We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
\begin{figure}[h]
\centering
\includegraphics[width=25pc]{../figs/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last thirty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

@@

The PPE is likewise useful for determining which parameters influence which model processes and where. For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}). Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome. The most influential parameters globally include parameters from each of these three biomes. We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).

@@

In this project, we have identified and perturbed 188 CLM parameters to create a large one-at-a-time parameter perturbation ensemble. This ensemble is useful for understanding parametric controls on CLM processes and has already been leveraged as a starting point for parameter optimization efforts. We made significant investments in software infrastructure to enable this experiment and reduce the burden of generating future PPEs.

@@

We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:violins}). That being said, the majority of parameter perturbations had very small effect, leading to output distributions strongly concentrated around the default simulation with long tails representing a small subset of parameters. In the case of gross primary production, only xx parameters had a parameter effect of more than 2\% under the control forcing scenario (Supp Figure \ref{supp:gpp}). As such, we would expect that the dimensionality of the effective parameter space for any given process would be significantly smaller than implied by the 188 parameters.

@@

A one-at-a-time PPE cannot capture parameter interactions, and our min/max sampling protocol precludes diagnosing non-linearities. As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty. The advantage of our dataset is that we can diagnose parameter effects without the uncertainty of an approximating emulator or regression model. As such, it is easy to diagnose which parameters are most influential on a given process, which has been leveraged for parameter filtering. We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation. Now in addition to seeing the definition of a given parameter, and the relevant equations, a model user can see the magnitude and spatial patterns of its effects (e.g. Figure \ref{fig:panel}).

We have already begun a follow-on experiment that perturbs a subset of parameters with the Latin hypercube sampling strategy to work towards calibrating leaf area index. By investing in the infrastructure underlying this first experiment, all of our subsequent experiments have required much less time and effort. We expect to continue to extend our efforts in this domain, as well as repeat these types of experiments with subsequent model releases.
